---
title: "Data Communication"
author: "MACSS 30500 <br /> University of Chicago"
output:
  xaringan::moon_reader:
    css: ["../cfss.css"]
    nature:
      highlightStyle: magula
      highlightLines: true
      highlightLanguage: r
      ratio: 16:9
      countIncrementalSlides: false
---

```{r setup, include = FALSE}
# generate CSS file
library(xaringanthemer)

#source(here::here("R", "slide-opts.R"))

# enable panelsets and default theme
xaringanExtra::use_panelset()
```

```{r pkgs, include = FALSE, cache = FALSE}
library(tidyverse)
library(palmerpenguins)
library(c3s2)
library(knitr)
library(patchwork)
library(flipbookr)

set.seed(1234)
theme_set(theme_gray(base_size = 16))
```

# Agenda

* Review layers + EDA
* Communicating with Data
  * Labels
  * Annotations
  * Scales
  * Themes
  * Layout
  

---
class: inverse, middle

# Communicating with data


---
### Setup

For this, we're going to work with these packages -- remember to install them if you haven't yet!

```{r, echo = T}
library(tidyverse)
library(scales)
library(ggrepel)
library(patchwork)
```

---

## Labels
There are many ways to label your axes -- two primary ways: 

* `labs()`: can specify x and y [labs example](https://r4ds.hadley.nz/communication.html#labels)
* `theme()`: can customize with more detail [theme example](https://ggplot2-book.org/themes.html)


--

We will mostly use labs in our examples (for now ... theme is particularly helpful for rotating axis text!)

---

### Labels: examples

Here are our baseline options:

```{r, echo = T, eval = F}
labs(
  x = "",
  y = "", 
  title = waiver(),
  subtitle = waiver(),
  caption = waiver(),
  tag = waiver(),
  alt = waiver(),
  alt_insight = waiver()
)
```

---
## Example: `taylor` package: 
Add labels to this plot:

```{r, echo = T}
library(taylor)
data(eras_tour_surprise) 
ggplot(eras_tour_surprise, aes(x = dress)) + 
  geom_bar(stat = "count", fill = c("deepskyblue", "blue", "deeppink", "darkgreen", "tomato1", 
                                      "steelblue1", "tomato4", "hotpink2", "chocolate3", "slateblue4", "gold2" ))
```
## Example: `taylor` package: 
Add labels to this plot:

```{r, echo = F}
library(taylor)
data(eras_tour_surprise) 
ggplot(eras_tour_surprise, aes(x = dress)) + 
  geom_bar(stat = "count", fill = c("deepskyblue", "blue", "deeppink", "darkgreen", "tomato1", 
                                      "steelblue1", "tomato4", "hotpink2", "chocolate3", "slateblue4", "gold2" )) +
  labs( x= "Dress Color", y = "Number of times worn", title = "Distribution of dresses worn during Eras Tour Surprise Songs", caption = "Source: `taylor` package"
  )
```

---

### Info/guidance on titles

* Book suggests only informative titles: re what is occurring
* I disagree
* Context and role: what is your goal and who is your audience

---
## Annotatations
Sometimes you want to add additional elements to your plots!

* `geom_text()`
* `geom_label_repel()`
* `geom_label()`
*  lines: `geom_hline()` and `geom_vline()`
*  shapes: `geom_rect()` and `geom_segment()`
* additional layers: e.g. a second dataset with `geom_point()`


---
## Annotatations


---
## Scales

---
## Themes
PREPARE FOR FUN!!!!

--

This is how you can make your best plot! The possibilities are endless!!
---
## Layout

---
# Recap:

* Labels
* Annotations
* Scales
* Themes
* Layout
  
