---
title: "Tidying Data: Practice in R"
author: "Jean Clipperton"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

```{r setup, include=FALSE}
library(prettydoc)
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

# Tidy practice
For this, we're going to begin with [exercises from the book](https://r4ds.hadley.nz/data-tidy.html). The second portion will involve a novel exercise. As always, we'll continue practicing cloning, staging, committing, and pushing back to GitHub!

NOTE: THIS WILL NOT COMPILE YET! There are blanks in here for you to fill in! You **CAN** move chunk-by-chunk to test things. Good luck!

## Starting data: Billboard

```{r}
data(billboard)

glimpse(billboard)

names(billboard)
```



## `Pivot_longer`
We'll continue with the Billboard data from Tidyverse:

```{r}
billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  )  %>% 
  mutate(
    week = parse_number(week)
  )

```

### Follow up

What would the following code do and why would we (not) use it? `drop_na(rank)`? How could we incorporate this in our code? 

### Challenge Q
Who had the most weeks in the dataset? You need to BOTH answer the question AND explain your process using FILTER. 

```{r}
billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )  %>%
  mutate(
    week = parse_number(week)
  )

 %>% filter(week == max(week))
```

### Next: who got stuck?
Note: this is tricky -- there are many ways to think about this, but are there any artists who charted only at one value for their rank? If so, who and how many weeks did they spend there? 

```{r}
bb %>% group_by(artist) %>% filter(() &()) %>% filter()
```


## `Pivot_wider`

From the book, we will use the `cms_patient_experience` dataset. 

```{r}
glimpse(cms_patient_experience)
```
Ugh, gross. We know we want one row per obs. Probably need org as our rows. 


```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = starts_with(""),
    names_from = ,
    values_from = 
  )
```

Now, what if we wanted our measure_cd variable as the observations? (you might do this for summarizing...we can discuss if this is `tidy` or not!)
```{r}
cms_patient_experience |> 
  pivot_wider(
    id_cols = measure_title,
    names_from = org_nm,
    values_from = 
  )
```



## Why not both kinds of pivots?
When sometimes you need both!

# Bringing it all together:

In this exercise, you will:

1. Pivot longer
1. Pivot wider
1. Write your file to a csv!

## Billboard: returns
Sometimes it can be easier to get your data into a long shape so you can nicely add or count within a group. you can then either pivot it back wider or

```{r}
billboard
```

Let's make some max/min columns and then put the data back -- consider this vs the following code block. 

```{r}
billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = 
  )  %>%
  mutate(
    week = parse_number(week)
  ) %>%
  group_by(track) %>%
  mutate(
    max_track_week = max(week),
    lowest_track_rank = min(rank)
  ) %>% 
  group_by(artist) %>%
  mutate(
    max_artist_week = max(week),
    lowest_artist_rank = min(rank)
  ) %>%
  pivot_wider(
    id_cols = c(track, artist, max_track_week, lowest_track_rank, max_artist_week, lowest_artist_rank),
    names_from = ,
    values_from = 
  )
```

```{r}
billboard %>% 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    values_drop_na = TRUE
  )  %>%
  mutate(
    week = parse_number(week)
  ) %>%
  group_by(track) %>%
  mutate(
    max_track_week = max(week),
    lowest_track_rank = min(rank)
  ) %>% 
  group_by(artist) %>%
  mutate(
    max_artist_week = max(week),
    lowest_artist_rank = min(rank)
  ) %>%
  select(!c(week, rank)) %>%
  unique()

```

## Wrap up
Now, let's export our data! (notice how we are using our relative file paths instead of absolute file paths!)

```{r}
write.csv(bat, "data/billboard_stats.csv")
```

